---
title       : Analysis of Variance (ANOVA)
subtitle    : Chapter 24
author      : Jason Bryer
job         : epsy530.bryer.org
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
--- &twocol
```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
require(ggplot2)
require(gdata)
require(psych)
options(digits=4, width=95)
opts_chunk$set(fig.path='Figures/Class24-')
opts_chunk$set(comment=NA)

require(granova)
require(granovaGG)
require(lattice)
data(singer)
data(rat)
hand <- read.csv('../Data/Textbook/Chapter_24/Hand_washing.csv')
```

## Analysis of Variance (ANOVA)

The goal of ANOVA is to test whether there is a discernible difference between the means of several groups.

#### Example

Is there a difference between washing hands with:  water only, regular soap, antibacterial soap (ABS), and antibacterial spray (AS)?
* Each tested with 8 replications
* Treatments randomly assigned

For ANOVA:
* The means all differ.
* Is this just natural variability?
* Null hypothesis:  All themeans are the same.
* Alternative hypothesis:  The means are not all the same.


--- &twocol

## Hand Washing Comparison

*** left

```{r hand-boxplot, fig.width=5, fig.height=5, tidy=FALSE}
ggplot(hand, aes(x=Method, 
	y=Bacterial.Counts)) + 
	geom_boxplot()
```

*** right

```{r, tidy=FALSE}
desc <- describeBy(hand$Bacterial.Counts, 
    hand$Method, mat=TRUE)[,c(2,4,5,6)]
desc$Var <- desc$sd^2
print(desc, row.names=FALSE)
```

---

## Washing type all the same?

* $ H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4 $
* By Central Limit Theorem:   
$$ Var(\bar{y}) = \frac{\sigma^2}{n} = \frac{\sigma^2}{8} $$
* Variance of {37.5, 92.5, 106.0, 117.0} is 1245.08.
* $\frac{\sigma^2}{8} = 1245.08$
* $\sigma^2 = 9960.64$
* This estimate for $\sigma^2$ is called the Treatment Mean Square, Between Mean Square, or $MS_T$
* Is this very high compared to what we would expect?


---

## How can we decide what $\sigma^2$ should be?

* Assume each washing method has the same variance.
* Then we can pool them all together to get the pooled variance ${ s }_{ p }^{ 2 }$
* Since the sample sizes are all equal, we can average the four variances: ${ s }_{ p }^{ 2 } = 1410.10$
* Other names for ${ s }_{ p }^{ 2 }$:  Error Mean Square, Within Mean Square, $MS_E$.

---

## Comparing $MS_T$ (between) and $MS_E$ (within)

$MS_T$
* Estimates $s^2$ if $H_0$ is true
* Should be larger than $s^2$ if $H_0$ is false

$MS_E$
* Estimates $s^2$ whether $H_0$ is true or not
* If $H_0$ is true, both close to $s^2$, so $MS_T$ is close to $MS_E$

Comparing
* If $H_0$ is true, $\frac{MS_T}{MS_E}$ should be close to 1
* If $H_0$ is false, $\frac{MS_T}{MS_E}$ tends to be > 1

--- &twocol

*** left

## The F-Distribution

* How do we tell whether $\frac{MS_T}{MS_E}$ is larger enough to not be due just to random chance
* $\frac{MS_T}{MS_E}$ follows the F-Distribution
	* Numerator df:  k – 1 (k = number of groups)
	* Denominator df:  k(n – 1)  
	* n = # observations in each group
* $F = \frac{MS_T}{MS_E}$ is called the F-Statistic.

A Shiny App by Dr. Dudek to explore the F-Distribution: <a href='http://shiny.albany.edu/stat/fdist/' window='_new'>http://shiny.albany.edu/stat/fdist/</a>

*** right

```{r fdistribution, fig.width=7, fig.height=5, tidy=FALSE}
df.numerator <- 4 - 1
df.denominator <- 4 * (8 - 1)
plot(function(x)(df(x,df1=df.numerator,
					df2=df.denominator)),
	 xlim=c(0,5), xlab='x', ylab='f(x)',
	 main='F-Distribution')
```

--- &twocol


## Back to Bacteria

*** left

* $MS_T = 9960.64$
* $MS_E = 1410.14$
* Numerator df = 4 – 1 = 3
* Denominator df = 4(8 – 1) = 28.

*** right

```{r}
(f.stat <- 9960.64 / 1410.14)
1 - pf(f.stat, 3, 28)
```

#### P-value for $F_{3,28} = 0.0011$ ###

---

## Assumptions and Conditions

* To check the assumptions and conditions for ANOVA, always look at  the side-by-side boxplots.
	* Check for outliers within any group.
	* Check for similar spreads.
	* Look for skewness.
	* Consider re-expressing.
* Independence Assumption
	* Groups must be independent of each other.
	* Data within each group must be independent.
	* Randomization Condition
* Equal Variance Assumption
	* In ANOVA, we pool the variances.  This requires equal variances from each group:  Similar Spread Condition.


---

## ANOVA in R

```{r}
aov.out <- aov(Bacterial.Counts ~ Method, data=hand)
summary(aov.out)
```

--- 

## Graphical ANOVA

```{r hand-granova, echo=TRUE, fig.width=7, fig.height=6}
hand.anova <- granova.1w(hand$Bacterial.Counts, group=hand$Method)
```


---

## Graphical ANOVA

```{r}
hand.anova
```

--- &twocol

## What Next?

*** left

ANOVA:  
* P-value large → Nothing left to say
* P-value small → Which means are large and which means are small?
* We can perform a t-test to compare two of them.
* We assumed the standard deviations are all equal.
* Use $s_p$, for pooled standard deviations.
* Use the Students t-model, df = N – k.

*** right

* If we wanted to do a t-test for each pair:
	* P(Type I Error) = 0.05 for each test.
	* Good chance at least one will have a Type I error.
* Bonferroni to the rescue!  
	* Adjust a to $\alpha/J$ where J is the number of comparisons.
	* 95% confidence (1 – 0.05) with 3 comparisons adjusts to (1 – 0.05/3) ≈ 0.98333.
	* Use this adjusted value to find t**.

